---
title: "Example 5-iris data"
author: "Pushpakanthie Wijekoon"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: readable 
    highlight: monochrome
---

## Some applications in Discriminant Analysis
1. Predicting success or failure of new products  
2. Accepting or rejecting admission to an applicant.  
3. Predicting credit risk category for a person  
4. Classifying patients into different categories.  

Themes for your R Markdown document can be found [here](https://www.datadreaming.org/post/r-markdown-theme-gallery/)
  
## Iris Data
The biologist, Edgar Anderson (Anderson 1935) collected data on three species of [iris flowers](https://drive.google.com/file/d/1cWlIYAqdsRYsgQU_zGn-a8v1uJl7j5ew/view?usp=sharing) found on the Gaspe Penninsula of Quebec, Canada. The species are “Setosa”, “Versicolor”, and “Virginica”. As a plant biologist, he wanted to devise a rule or test to distinguish one species from another. To find an answer for this question, he carefully measured the length and width of two parts of each flower: the flower petals and sepals (the green leaves that enclose the flower).    
The variables in this data set are    
1. Sepal.Length in cm  
2. Sepal.Width in cm  
3. Petal.Length in cm  
4. Petal.Width in cm  
5. Species: setosa, versicolour, virginica  
  
The famous statistician, R. A. Fisher (Fisher 1936) used this data to answer Anderson’s questions, and he introduced the method of discriminant analysis.  
  
We can access this dataset from the "datasets" package.    

## Load iris data

```{r}
library(datasets)
str(iris)
head(iris)
write.csv(iris, file = "irisdata.csv")
```
## Correlation plots
Another way of getting pairs plot with the relevant correlations.  

```{r}
library(psych)
pairs.panels(iris[1:4],
             gap = 0,
             bg = c("red", "green", "blue")[iris$Species],
             pch = 21)
```

### Performing LDA 
First, we partition the data set to get training and testing data. We use 75% of the data for training data set.      
```{r}
#random partition, 75% of the sample size
smp_size = floor(0.75 * nrow(iris))
## set the seed to make your partition reproducible
set.seed(1337)
train_ind = sample(seq_len(nrow(iris)), size = smp_size)
train = iris[train_ind, ]
test = iris[-train_ind, ]
```

The MASS package contains the function *lda()* for performing linear discriminant analysis.  
The default option of lda() function is CV=TRUE, which indicates that the within class covariance are equal.    

```{r}
library(MASS)
iris.lda <- lda(Species ~ ., data = train)
iris.lda
```
The interpretation of the discriminant function can be made in several ways. The most simple method is to rank the absolute value of the coefficients, and determine contribution based on the order of the coefficients. Another method is to perform a partial F-test to find the significance of the variables.  
Since we have three groups in this example, we have two discriminant functions, LD1 and LD2.    
LD1: $y_{1}=0.8284Sepal.Length+1.34399Sepal.Width-2.1418Petal.Length-3.0875Petal.Width$   
   
LD2: $y_{2}=-0.2275Sepal.Length-2.14496Sepal.Width+1.2461Petal.Length-3.2628Petal.Width$  

Percentage separations achieved by the first discriminant function is 98.9% and second is 1.1%.  

What are the most significant measurements?    

## Model Accuracy
Then, we find the model accuracy.  

```{r}
#Check the accuracy of our analysis
Predictions=predict(iris.lda,test)
table(Predictions$class, test$Species, dnn = c('Actual Group','Predicted Group'))

# Model accuracy and Predictions
mean(Predictions$class == test$Species)

```

Model accuracy is about 95%, and it is a very high value. Two virginica species are misspecified as versicolor.  
Using the discriminant functions, now we predict the total data set. 

```{r}
# Prediction of total data
Predict_tot=predict(iris.lda,iris)
table(Predict_tot$class, iris$Species, dnn = c('Actual Group','Predicted Group'))
```

## Plot the linear discriminants visually
### Stacked histograms
As a final step, we will plot the linear discriminants and visually see the differentition ability. The ldahist() function helps make the separator plot. For the data into the ldahist() function, we can use the x[,1] for the first linear discriminant and x[,2] for the second linear discriminant and so on.  

```{r}
#Plot the predictions - first linear discriminant
ldahist(data = Predict_tot$x[,1], g=iris$Species)
```

The observations are almost completely separated by the first linear discriminant. Now, we check the discriminant ability of the second linear discriminant.  

```{r}
#Plot the predictions - second linear discriminant
ldahist(data = Predict_tot$x[,2], g=iris$Species)
```

The above plots of the second linear discriminant, it is clear that we can hardly differentiate between the three groups. Therefore, the  first discriminat function is sufficient to sepearte the species.    

### Biplot
Biplot based on LD1  and LD2 is given below. Setosa is separated very clearly, and some overlapping can be observed between Versicolor and virginica.    

Based on arrows, Sepal width and sepal length explained more for setosa, petal width and petal length explained more for versicolor and virginica.  

```{r}
#install.packages('devtools')
#library(devtools)
#install_github('fawda123/ggord')
library(ggord)
ggord(iris.lda, train$Species, ylim = c(-10, 10))
```

## Performing Quadratic Discriminant Analysis (QDA)
The qda() function in MASS package can be used for QDA.  

```{r}
#QDA
qda_iris=qda(Species~.,data=train)
qda_iris
 
```


```{r}
#Check the accuracy of our analysis of qda
Predictions_qda=predict(qda_iris,test)
table(Predictions_qda$class, test$Species, dnn = c('Actual Group','Predicted Group'))

# Model accuracy
mean(Predictions_qda$class == test$Species)
```

Since the data has a linear relation, the qda function also applies the same statistics and returns similar results.  

The accuracy of QDA (0.9473684) is also similar to the accuracy of LDA (0.9473684). Therefore, LDA is recommended to use since tit is he simple method.    

## Comparison of Partition plot for LDA and QDA
Partition plot provides the classification of each and every combination in the training dataset.  

```{r}
library(klaR)
partimat(Species~., data = train, method = "lda")
```

```{r}
partimat(Species~., data = train, method = "qda")
```
